---
title: "Plot xenium core spatial expression"
author: "User"
date: "`r Sys.Date()`"
params:
  data_path: ""
  out_path: ""
  config_path: ""
output: 
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 2
    number_sections: true
    theme: flatly
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

# Document contents
The objective of this document is to plot spatial expression in data from the G34 project processed with Xenium.

# Extract parameters and read config
In this section we extract the paths to the data and desired output directories and the config.
```{r}
# Get paths passed from SLURM
data_path    <- params$data_path
out_path  <- params$out_path
config_path <- params$config_path

config <- read.csv(config_path)
```


# Load libraries
```{r}
library(dplyr)
library(Seurat)
library(tibble)
library(ggplot2)
library(colorRamp2)
library(RColorBrewer)
library(cowplot)
library(arrow)
library(stringr)
library(readr)
library(purrr)
library(here)
library(glue)
library(terra)
library(patchwork)
library(scales)
```


# Generate plots
In this section, we generate the plots for each row in the config. This is done through a for loop that extracts the following information for each query:

- Sample name
- TMA name
- Features to visualize (either genes or meta data columns)
- Visualization coordinates (either full core or crop)
- Normalization (either raw or log normalized counts)

If the user wants to visualize the feature across the full core, the spatial plot will be generated showing the locations of cells (depicted as points) colored by the intensity of the feature being visualized. If the user wants to visualize a crop, the coordinates of the crop provided will be used to show the spatial distribution of a spatial subset of the data, where the cells will be represented by their segmentation polygons. For crops, the IF images with and without the segmentation boundaries will be provided by default.
```{r}

# Loop through all rows of the config and extract meta data
for (i in 1:nrow(config)) {
  # Extract row in config being processed
  row <- config %>% filter(row_number() == i)
  
  # Extract parameters
  sample_name <- row %>% pull(sample_name)
  tma_name <- row %>% pull(tma_name)
  features <- row %>% pull(feature) %>% str_split(";") %>% .[[1]] %>% trimws()
  full_core <- row %>% pull(full_core)
  crops <- row %>% pull(crops)
  xmin <- row %>% pull(xmin)
  xmax <- row %>% pull(xmax)
  ymin <- row %>% pull(ymin)
  ymax <- row %>% pull(ymax)
  
  # Create path to save outputs
  save_path <- glue("{out_path}/{tma_name}/{sample_name}")
  dir.create(save_path, recursive = T)
  
  # Get data
  seurat <- readRDS(glue("{data_path}/seurat_objects/{tma_name}/{sample_name}.rds"))
  
  # Generate plots for the full core
  if (full_core == "Y") {
  plots <- list()
    
    for (f in features) {
      current_f <- f 
      
      # If feature is a gene then extract raw counts
      if (f %in% rownames(seurat)) {
        # Extract raw counts
        exp <- FetchData(object = seurat, vars = f, layer = "counts")
        f_to_plot <- paste0(f, "_raw_full") # Unique name
        colnames(exp) <- f_to_plot
        seurat <- AddMetaData(object = seurat, metadata = exp)
        
        # Visualize expression
        p <- ggplot(seurat@meta.data, aes(x = x, y = y, color = .data[[f_to_plot]])) +
        ggrastr::geom_point_rast(size = 0.08, raster.dpi = 300) +
        scale_color_distiller(palette = "Blues", 
                              direction = 1, 
                              limits = c(0, 
                                         quantile(seurat@meta.data[[f_to_plot]], 0.99, na.rm = TRUE)), 
                              oob = scales::squish) +
        coord_fixed() + 
        theme(
          aspect.ratio = 1,
          axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          legend.position = "bottom",
          axis.title = element_blank(),
          panel.border = element_rect(colour = "grey80", fill = NA, linewidth = 1),
          panel.background = element_blank(),
          panel.grid = element_blank(),
          plot.title = element_text(hjust = 0.5)
        ) +
        ggtitle(sample_name) + labs(color = paste0(current_f, " expression"))
        
        plots[[f]] <- p
        
      } else {
        f_to_plot <- f # For clusters/categorical metadata
      }
    }
  
    # Get plots for QC metrics
  p_counts <- ggplot(seurat@meta.data, aes(x = x, y = y, color = .data[["nCount_Xenium"]])) +
        ggrastr::geom_point_rast(size = 0.08, raster.dpi = 300) +
        scale_color_distiller(palette = "Blues", 
                              direction = 1, 
                              limits = c(0, 
                                         quantile(seurat@meta.data[["nCount_Xenium"]], 0.99, na.rm = TRUE)), 
                              oob = scales::squish) +
        coord_fixed() + 
        theme(
          aspect.ratio = 1,
          axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          legend.position = "bottom",
          axis.title = element_blank(),
          panel.border = element_rect(colour = "grey80", fill = NA, linewidth = 1),
          panel.background = element_blank(),
          panel.grid = element_blank(),
          plot.title = element_text(hjust = 0.5)
        ) +
        ggtitle(sample_name) + labs(color = "UMIs/cell")
  
  p_features <- ggplot(seurat@meta.data, aes(x = x, y = y, color = .data[["nFeature_Xenium"]])) +
        ggrastr::geom_point_rast(size = 0.08, raster.dpi = 300) +
        scale_color_distiller(palette = "Blues", 
                              direction = 1, 
                              limits = c(0, 
                                         quantile(seurat@meta.data[["nFeature_Xenium"]], 0.99, na.rm = TRUE)), 
                              oob = scales::squish) +
        coord_fixed() + 
        theme(
          aspect.ratio = 1,
          axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          legend.position = "bottom",
          axis.title = element_blank(),
          panel.border = element_rect(colour = "grey80", fill = NA, linewidth = 1),
          panel.background = element_blank(),
          panel.grid = element_blank(),
          plot.title = element_text(hjust = 0.5)
        ) +
        ggtitle(sample_name) + labs(color = "genes/cell")
  
    # Save plots as a pdf
    pdf(file = glue("{save_path}/full_core_visualizations_{i}.pdf"))
    for (f_name in names(plots)) {
      p <- plots[[f_name]]
      print(p)
    }
    print(p_counts)
    print(p_features)
    dev.off()
  }
  
  # Generate plots for crops
  if (crops == "Y") {
    # Initialize empty list for plots
    feature_plots <- list()
    
    # Load polygons
    polygons <- read_parquet(glue("{data_path}/polygons/{tma_name}/cell_boundaries.parquet"))
    
    # Loop through vector of features
    for (f in features) {
      current_f <- f # Keep the original name for labeling plot
      
      # If feature is a gene, extract counts 
      if (f %in% rownames(seurat)) {
        # Pull raw counts from counts slot
        exp <- FetchData(object = seurat, vars = f, layer = "counts")
        colnames(exp) <- paste0(f, "_raw_plot")
        seurat <- AddMetaData(object = seurat, metadata = exp)
        f_to_plot <- paste0(f, "_raw_plot")
        
        # Generate plot
        p <- polygons %>%
        filter(vertex_x > xmin & vertex_x < xmax & vertex_y > ymin & vertex_y < ymax) %>%
        left_join(seurat@meta.data %>% 
                  select(all_of(f_to_plot)) %>% 
                  mutate(cell_id = rownames(.)), by = "cell_id") %>% 
        ggplot(mapping = aes(x = vertex_x, y = vertex_y, group = cell_id)) +
        geom_polygon(aes(fill = .data[[f_to_plot]]), 
                     colour = "black", 
                     linewidth = 0.1) +
        scale_fill_viridis_c(
          option = "magma", 
          direction = 1, 
          limits = c(0, quantile(seurat@meta.data[[f_to_plot]], 0.99, na.rm = TRUE)), 
          oob = scales::squish,
          na.value = "grey90") +
        coord_fixed() + 
        theme_void() +
        theme(
          panel.border = element_rect(colour = "grey80", fill = NA, linewidth = 1),
          plot.title = element_text(hjust = 0.5, size = 24),
          legend.position = "bottom") + 
        ggtitle(sample_name) + labs(fill = paste0(current_f, " expression"))
      
      feature_plots[[f]] <- p
      
      } else {
        f_to_plot <- f # For any other meta data
      }
    }
    
       # Get plots for QC metrics
  p_counts <- polygons %>%
        filter(vertex_x > xmin & vertex_x < xmax & vertex_y > ymin & vertex_y < ymax) %>%
        left_join(seurat@meta.data %>% 
                  select(all_of("nCount_Xenium")) %>% 
                  mutate(cell_id = rownames(.)), by = "cell_id") %>% 
        ggplot(mapping = aes(x = vertex_x, y = vertex_y, group = cell_id)) +
        geom_polygon(aes(fill = .data[["nCount_Xenium"]]), 
                     colour = "black", 
                     linewidth = 0.1) +
        scale_fill_viridis_c(
          option = "magma", 
          direction = 1, 
          limits = c(0, quantile(seurat@meta.data[["nCount_Xenium"]], 0.99, na.rm = TRUE)), 
          oob = scales::squish,
          na.value = "grey90") +
        coord_fixed() + 
        theme_void() +
        theme(
          panel.border = element_rect(colour = "grey80", fill = NA, linewidth = 1),
          plot.title = element_text(hjust = 0.5, size = 24),
          legend.position = "bottom") + 
        ggtitle(sample_name) + labs(fill = "UMIs/cell")
  
  p_features <- polygons %>%
        filter(vertex_x > xmin & vertex_x < xmax & vertex_y > ymin & vertex_y < ymax) %>%
        left_join(seurat@meta.data %>% 
                  select(all_of("nFeature_Xenium")) %>% 
                  mutate(cell_id = rownames(.)), by = "cell_id") %>% 
        ggplot(mapping = aes(x = vertex_x, y = vertex_y, group = cell_id)) +
        geom_polygon(aes(fill = .data[["nFeature_Xenium"]]), 
                     colour = "black", 
                     linewidth = 0.1) +
        scale_fill_viridis_c(
          option = "magma", 
          direction = 1, 
          limits = c(0, quantile(seurat@meta.data[["nFeature_Xenium"]], 0.99, na.rm = TRUE)), 
          oob = scales::squish,
          na.value = "grey90") +
        coord_fixed() + 
        theme_void() +
        theme(
          panel.border = element_rect(colour = "grey80", fill = NA, linewidth = 1),
          plot.title = element_text(hjust = 0.5, size = 24),
          legend.position = "bottom") + 
        ggtitle(sample_name) + labs(fill = "genes/cell")
    
    
    # Define function for extracting xenium images
    GetXeniumImage <- function(xmin, xmax, ymin, ymax, path_8bit, brighten = 1) {
      # Load image
      img <- terra::rast(path_8bit)
      
      # Transform micron to pixel coords
      res_um <- 0.2125
      img_width_um <- ncol(img) * res_um
      img_height_um <- nrow(img) * res_um
      terra::ext(img) <- c(0, img_width_um, 0, img_height_um)
      
      # 3. Safety Check: Ensure requested coordinates are within the image bounds
      # This prevents the "crop" function from returning an empty object
      xmin <- max(0, xmin)
      ymin <- max(0, ymin)
      xmax <- min(img_width_um, xmax)
      ymax <- min(img_height_um, ymax)
      
      if (xmin >= xmax || ymin >= ymax) {
        warning(paste("Invalid crop coordinates for", path_8bit, "- returning empty dataframe"))
        return(data.frame(x=numeric(), y=numeric(), intensity=numeric()))
      }
      
      # Define and extract crop
      e <- terra::ext(c(xmin, xmax, ymin, ymax))
      img_subset <- terra::crop(img, e)
      
      # Convert crop to data frame
      img_subset_df <- as.data.frame(img_subset, xy = TRUE)
      
      # Set column names
      colnames(img_subset_df) <- c("x", "y", "intensity")
      
      # Apply brightness for dim stains
      if (brighten != 1) {
        img_subset_df$intensity <- img_subset_df$intensity * brighten
        img_subset_df$intensity[img_subset_df$intensity > 255] <- 255
      }
      
      return(img_subset_df)
    }

    # Generate list of parameters for the IF plots
    stain_settings <- list(
      DAPI = list(img = "ch0000_dapi.tif", color = "#ade8f4", bright = 1.5),
      ATP_CD45_ECad = list(img = "ch0001_atp1a1_cd45_e-cadherin.tif", color = "#FF00FF", bright = 3),
      RNA18S = list(img = "ch0002_18s.tif", color = "#fb8500", bright = 3.5),
      aSMA_Vim = list(img = "ch0003_alphasma_vimentin.tif", color = "#00ff2b", bright = 2))
    
    # Define function to generate plots
    PlotIFStains <- function(stain_name, stain_params, add_polygons = FALSE) {
      
      # Load image 
      img <- GetXeniumImage(
        xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, 
        path_8bit = glue("{data_path}/images/{tma_name}/{stain_params$img}"), 
        brighten = stain_params$bright
      ) %>%
        filter(x > xmin & x < xmax & y > ymin & y < ymax)
    
      # Get base plot
      p <- ggplot() +
        geom_raster(data = img, aes(x = x, y = y, fill = intensity)) +
        scale_fill_gradient(low = "black", high = stain_params$color, guide = "none") +
        coord_fixed(expand = FALSE) +
        theme_void() + 
        theme(
          panel.border = element_rect(colour = "grey80", fill = NA, linewidth = 1),
          plot.title = element_text(hjust = 0.5, size = 24)
        ) +
        labs(title = stain_name)
    
      if (add_polygons) {
        p <- p + geom_polygon(
          data = polygons %>% filter(vertex_x > xmin & vertex_x < xmax & vertex_y > ymin & vertex_y < ymax), 
          aes(x = vertex_x, y = vertex_y, group = cell_id),
          fill = NA, colour = "yellow", linewidth = 0.2, alpha = 0.8
        )
      }
      
      return(p)
    }
    
    # Get stain plots with and without polygons
    stain_plots <- lapply(names(stain_settings), function(n) {
      PlotIFStains(stain_name = n, stain_params = stain_settings[[n]])
    })
    
    stain_polygons_plots <- lapply(names(stain_settings), function(n) {
      PlotIFStains(stain_name = n, stain_params = stain_settings[[n]], add_polygons = TRUE)
    })
    
    # Save plots as pdf
    pdf(file = glue("{save_path}/crop_visualizations_{i}.pdf"))

    # Use the list dedicated to the CROPPED ggplot objects
    for (f_name in names(feature_plots)) {  
      p <- feature_plots[[f_name]]
      print(p)
    }
    
    print(p_counts)
    print(p_features)

    # Save IF stains
    print(plot_grid(plotlist = stain_polygons_plots, nrow = 2, ncol = 2, align = "hv"))
    print(plot_grid(plotlist = stain_plots, nrow = 2, ncol = 2, align = "hv"))

    dev.off()
  }
}


```


# Reproducibility
```{r}
sessionInfo()
```

